Reconstructing the Lensing Mass in the Universe \\ from Photometric Catalog Data

### 2012-10-30 doc/paper-lcr.tex
intro done, moving on

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected by measuring the
small distortions it induces on the images of many background galaxies
in the field. \todo{PJM}{Write about Bernstein project in 0957.}

\todo{PJM}{Rearrange following section. After lunch.}

To date, attempts at reconstructing the local and line of sight mass
distribution in strong lens fields (what \citeauthor{WongEtal2011} call ``the
full environment'') have focused on photometric and  spectroscopic surveys to
detect galaxy groups and ascertain galaxies' membership in order to understand
the external shear apparently required by the strong lens models. Surveys by
\citet{Fassnacht+Lubin2002,AugerEtal2007,WilliamsEtal2006,MomchevaEtal2006}
all found groups of galaxies both hosting and lying  along the line of sight
to known gravitational lenses. Initial estimates of the effect of these
structures suggested the importance of assigning mass to both galaxy and group
halos, and to allowing for significant uncertainty in the centroids of group
halos. The probability density function (PDF) for the external convergence for
that lens, $\Pr(\xkappa)$, is the target for any reconstruction analysis, and
is the key quantity for any ensemble analysis, since by marginalising over it
we can correctly propagate the uncertainties due to line-of-sight perturbations. Coming
closest to estimating this quantity from survey data, \citet{WongEtal2011} ran
Monte Carlo reconstructions of the mass in nine lens fields based on
\citeauthor{WilliamsEtal2006} and \citeauthor{MomchevaEtal2006}'s
photometric and spectroscopic measurements of galaxies close to the line of
sight, and compared the resulting predicted shear distributions with the
external shear demanded by the strong lens model. Their reconstructions show most
of the shear being generated by bright galaxies within 2~arcminutes of the
lens, and that both line of sight structures and the group of galaxies in each
lens plane contribute significant proportions of the shear. They find
significant discrepancies between the lens model shear and the Monte Carlo
predictions, and are unable to distinguish between the environment modelling
and the strong lens modelling as the cause of these discrepancies.

These observational investigations have provided very useful insight into the
problem.  Coming from a more theoretical direction, large numerical
simulations have been used to predict, from first principles, the distribution
of likely external shears and convergences along strong lens lines of sight. 
\citet{Holder+Schechter2003} and \citet{Dalal+Watson2004}\todo{phil}{Dalal+Watson2004 doesn't seem to exist} carried out
ray-tracing calculations in N-body simulations to estimate the  distribution
of external shear values.  \citet{HilbertEtal2009} performed similar ray-tracing
experiments through the Millenium Simulation \citep{SpringelEtal2005},
generating a predicted $\kappa$ at every position in a simulated sky. \citet{HilbertEtal2009}
found that \MS lines of sight with strong lenses were not biased towards high $\kappa$,
although selection functions for discovering a strong lens were not taken into account.
The first step towards using this excellent calibration resource was taken by
\citet{SuyuEtal2010}, who selected Millenium Simulation lines of sight that
 had comparable apparent galaxy
overdensity in a 45 arcsec radius aperture down to $i < 24.5$. This overdensity
could be compared to the overdensity of the B1608$+$656 line of sight
\citep{FassnachtEtal2011}, and hence an estimate of  $\Pr(\kappa)$ could be
made. B1608$+$656 is indeed a high $\kappa$ line of sight, but it does not reside
in a typical lens environment.

In this paper, we bring these observational and theoretical approaches
together and  use numerical simulations to {\it test and calibrate} a
reconstruction of the  external convergence. Ray-tracing through the
simulations gives a ``truth'' that we can attempt to recover.  In this paper
we use the Millenium Simulation mock galaxy catalogs and ray-traced lensing
effects to test the accuracy of line of sight mass reconstructions.
Probabilistically assigning mass and redshift to every observed galaxy in the
field around a time delay lens, we generate Monte Carlo sample line of sight
mass distributions, and estimate the external convergence due to each one.
By using the Millenium siumlation as a calibration tool, we can convert the
total contribution from halos, $\Pr(\kappah)$, into a PDF for the likely convergence
along the line of sight $\Pr(\kappa)$.

We aim to answer the following questions:
\begin{itemize}
\item Faint galaxies, filaments and dark structures will not appear in a photometric
object catalog, but they will contribute convergence at some level. How much of
the total external convergence in a time delay lens system comes from visible
galaxies? How does this change as a function of magnitude cut? 
\item Can the true (ray-traced) convergence be recovered from a calibrated halo model
reconstruction? What scatter and residual bias are induced by the reconstruction
process? Which aspects of the model dominate these uncertainties? 
\item When faced with a newly-detected lens, surrounded by galaxies on the sky,
we have some choices to make when planning follow-up observations.
Which are the most important galaxies, with regard to the external convergence
produced? Can nearby groups and clusters be straightforwardly accounted for? If
lenses with such massive structures nearby are discarded, what impact does that
have on the distance accuracy?
\end{itemize}

This paper is organized as follows. In~\Sref{sec:theory} we review the relevant
gravitational lensing theory, before introducing our simple reconstruction 
model in \Sref{sec:model}.  We then test this model in two phases: first, in
\Sref{sec:knownMh+z}, with known redshift and halo mass for every galaxy in a
lens field, in order to quantify the irreducible uncertainty due to unseen mass,
and second, in \Sref{sec:obsMstar+z}, with realistic observational uncertainties
on the field galaxies' stellar masses and redshifts. In \Sref{sec:biases} we
investigate potential systematic errors induced by selecting a subset of sightlines.
We discuss our results in \Sref{sec:discuss} before
concluding in \Sref{sec:conclude}.

Throughout this paper magnitudes are given in the AB system and
we adopt the Millenium Simulation's ``concordance'' parameters for our reference cosmology, \ie
$h=0.73$, $\Omega_m=0.25$ and $\Omega_\Lambda=0.75$, where the symbols indicate
the Hubble Constant in units of 100 km s$^{-1}$ Mpc$^{-1}$ and the matter and
dark energy density of the Universe in units of the critical density
.%\citep[e.g.\ ][]{KomEtal09}.

>>>>

Several attempts have been made to do better than simply constructing a
large sample of objects, and averaging over the resulting convergence
distribution. The weak lensing effect can be detected by measuring the
small distortions it induces on the images of many background galaxies
in the field \citep[see e.g.][for a review]{SaasFee2006}. In a 30''
radius aperture around the strong lens Q0957$+$561,
\citet{NakajimaEtal2009} used deep HST imaging to measure a weak lensing
convergence of $0.17/pm0.06$, which was used in the time delay distance
measurement of \citep{FadelyEtal2009}. While this system has an
unusually high external convergence due to its location on the edge of a
cluster of galaxies, the precision with which the convergence at the
lens position can be estimated is indicative of what can be achieved
with high quality weak lensing shear data. We note that this uncertainty
is not dependent on the value of the convergence, but instead depends
primarily on the number density of weakly lensed, measurable galaxies
available.

In order to make more precise estimates of the convergence at a given
sky position, we need more information.   There have been some attempts
to reconstruct the local and line of sight mass distribution in strong
lens fields in some detail, based on photometric and spectroscopic
surveys of the galaxies nearby. These efforts have focused on  detecting
galaxy groups, ascertaining the visible galaxies' membership of those
groups, and assigning mass to the group and galaxy halos in order to
predict the external shear apparently required by the strong lens
models. Surveys by
\citet{Fassnacht+Lubin2002,AugerEtal2007,WilliamsEtal2006,MomchevaEtal2006}
all found groups of galaxies both hosting, and lying  along the line of
sight to, known gravitational lenses. Initial estimates of the effect of
these structures suggested the importance of assigning mass to both
galaxy and group halos, and to allowing for significant uncertainty in
the centroids of group halos.

The probability density function (PDF) for the external convergence
given our available data $\data$, $\Pr(\xkappa|\data)$,  is the  key
quantity we seek, since by marginalising over this distribution we can
correctly propagate the uncertainties due to line-of-sight perturbations
(as \citeauthor{FadelyEtal2009} did).
\citet{WongEtal2011} came close to estimating this quantity from their
survey data, running Monte Carlo reconstructions of the mass in nine
lens fields based on \citeauthor{WilliamsEtal2006} and
\citeauthor{MomchevaEtal2006}'s photometric and spectroscopic
measurements of galaxies close to the line of sight, and comparing the
resulting predicted shear distributions with the external shear demanded
by the strong lens model. Their reconstructions showed most of the shear
to have been generated by bright galaxies within 2~arcminutes of the
lens, and that both line of sight structures and the group of galaxies
in each lens plane contribute significant proportions of the shear. They
found significant discrepancies between the lens model shear and the
Monte Carlo predictions, but were unable to distinguish between the
environment modelling and the strong lens modelling as the cause of
these discrepancies.

These observational investigations have provided some very useful
insight into the problem of convergence estimation.  Coming from a more
theoretical direction, large numerical simulations have been used to
predict, from first principles, the distribution of likely external
shears and convergences along various lines of sight, including those
containing strong lens systems.  \citet{Holder+Schechter2003} and
\citet{DalalEtal2005} carried out ray-tracing calculations in N-body
simulations to estimate the distribution of external shear values. 
\citet{HilbertEtal2009} performed similar ray-tracing experiments
through the Millenium Simulation \citep{SpringelEtal2005}, generating a
predicted $\kappa$ at every position in a simulated sky.
\citet{HilbertEtal2009} found that \MS lines of sight with strong lenses
were not biased towards high $\kappa$, although selection functions for
discovering or using particular samples of strong lenses were not taken
into account. The first step towards using this excellent calibration
resource was taken by \citet{SuyuEtal2010}, who selected \MS lines of
sight by their apparent galaxy overdensity, in a 45 arcsec radius
aperture down to $i < 24.5$, to match the observed overdensity towards
the time delay lens B1608$+$656 \citep{FassnachtEtal2011}. The resulting
distribution of $\kappa$ values from the ray-tracing was taken to be an
estimate of  $\Pr(\kappa)$. B1608$+$656 is indeed a high $\kappa$ line
of sight, but it does not reside in a massive cluster environment. In a
companion paper to this one, \citet{GreeneEtal2012} investigated
improvements to this method by weighting the galaxy counts by galaxy
mass, and perpendicular distance to the line of sight, again using the
\MS mock catalogs and their associated $\kappa$ values to construct
$\Pr(\kappa|\data)$.

In this paper, we combine the halo model reconstruction approach of
\citeauthor{WongEtal} and others with the idea of calibrating to
simulations from \citeauthor{SuyuEtal2009} in order  to build a
framework for bringing all available information to bear on the problem
of estimating the weak the lensing convergence due to mass along the
line of sight to a distant source.  Taking the $\kappa$ values from
\citeauthor{HilbertEtal2009}'s ray-tracing calculations as the ``truth''
that we have to recover, we use the \MS mock galaxy catalogs to test the
accuracy of line of sight mass reconstruction. Probabilistically
assigning mass and redshift to every observed galaxy in a given observed
field, we generate Monte Carlo sample line of sight mass distributions,
and so construct the PDF $\Pr(\kappa|\data)$.  We aim to answer the
following questions:

\item Faint galaxies, filaments and dark structures will not appear in
any photometric object catalog, but they will contribute convergence at
some level. How much of the total convergence comes from visible
galaxies, and how much effect do dark structures and voids have?

\item Can the true convergence be recovered from a calibrated halo model
reconstruction? What scatter and residual bias are induced by the
reconstruction process, and how might these be reduced in the future?

\item We have some choices to make when planning follow-up observations
of interesting sightlines. How should we select objects to achieve the
highest accuracy convergence reconstruction, and hence the highest
accuracy global parameter estimates?

This paper is organized as follows. We review the relevant gravitational
lensing theory in~\Sref{sec:theory}, and the Millenium Simulation
ray-traced convergences and mock catalogs in \Sref{sec:MS}, before
introducing our simple reconstruction  model in \Sref{sec:model}. We
then test this model in two phases: first, in \Sref{sec:knownMh+z}, with
known redshift and halo mass for every galaxy in a lens field, in order
to quantify the irreducible uncertainty due to unseen mass, and second,
in \Sref{sec:obsMstar+z}, with realistic observational uncertainties on
the observed galaxies' stellar masses and redshifts. In
\Sref{sec:biases} we investigate the potential systematic error induced
by selecting a subset of sightlines. We discuss our results in
\Sref{sec:discuss} before concluding in \Sref{sec:conclude}.

Throughout this paper magnitudes are given in the AB system and
we adopt the Millenium Simulation's ``concordance'' parameters for our reference cosmology, \ie
$h=0.73$, $\Omega_m=0.25$ and $\Omega_\Lambda=0.75$, where the symbols indicate
the Hubble Constant in units of 100 km s$^{-1}$ Mpc$^{-1}$ and the matter and
dark energy density of the Universe in units of the critical density.
.%\citep[e.g.\ ][]{KomEtal09}.

### 2012-10-30 doc/references.tex
working on intro
intro done, moving on

\bibitem[Dalal(2005)]{Dalal2005} Dalal, N.\ 2005, 25 Years After 
the Discovery: Some Current Topics on Lensed QSOs,

@@

@@

@@

\bibitem[\protect\citeauthoryear{Linder 
\& Holz}{2004}]{Linder+Holz2004} Linder E.~V., Holz D.~E., 2004, AAS, 36, \#131.02

@@

@@

\bibitem[\protect\citeauthoryear{\tom{which paper}}{{9999}}]{xxx} 
None

>>>>

% \bibitem[Dalal(2005)]{Dalal2005} Dalal, N.\ 2005, 25 Years After 
% the Discovery: Some Current Topics on Lensed QSOs,

@@

\bibitem[Fadely et al.(2010)]{FadelyEtal2009} Fadely, R., Keeton, 
C.~R., Nakajima, R., \& Bernstein, G.~M.\ 2010, \apj, 711, 246

@@

\bibitem[Holz 
\& Linder(2005)]{Holz+Linder2005} Holz, D.~E., \& Linder, E.~V.\ 2005, \apj, 631, 678

@@

% \bibitem[\protect\citeauthoryear{Linder 
% \& Holz}{2004}]{Linder+Holz2004} Linder E.~V., Holz D.~E., 2004, AAS, 36, \#131.02

@@

\bibitem[Nakajima et al.(2009)]{NakajimaEtal2009} Nakajima, R., 
Bernstein, G.~M., Fadely, R., Keeton, C.~R., 
\& Schrabback, T.\ 2009, \apj, 697, 1793

@@

% \bibitem[\protect\citeauthoryear{\tom{which paper}}{{9999}}]{xxx} 
% None
